---
title: "ROH-hotspot identification"
output:
  html_document:
    toc: true
    toc_depth: 3  
date: "2024-10-06"
editor_options: 
  markdown: 
    wrap: 72
---
# 0: Preparation
Defining the input and output files
```{r}
# Clean the working environment
rm(list = ls())
knitr::opts_chunk$set(echo = TRUE)

################################### 
#Defining Input parameters
################################### 
roh_hotspot_threshold_percentage <- 0.01 # 0.01 represents top 1 % most homozygous region
max_hotspot_windows_gap <- 2*10^6 # ROH-hotspot windows that are less than 2Mb apart are merged together (see section 1.3)
####################################  
# Defining the input file
#################################### 
input_file <- Sys.getenv("input_bed_file")
input_file

simulation_processing <- as.logical(Sys.getenv("simulation_processing"))
simulation_processing

if (simulation_processing) { 
  max_length_chromosome <- as.numeric(Sys.getenv("max_length_chromosome"))
  max_length_chromosome
}
Sys.getenv()  #

## Extracting the prefix of the file used  
# Extracting the basename (filename without extension) from the file path
filename <- basename(input_file)
# Split the filename by "_ROH_freq.bed" and get the first part
prefix <- strsplit(filename, "_ROH_freq.bed")[[1]][1]
# Output the prefix
print(prefix)
chr_prefix <- "CHR"
#################################### 
# Defining the output dirs
#################################### 
# Define the output directory based on the variable passed from the Bash script
output_dir <- Sys.getenv("output_directory")
output_dir
# Output dir for section 3.2
output_dir_gapless_roh_hotspots <- Sys.getenv("gapless_roh_hotspots_directory")
output_dir_gapless_roh_hotspots
# Output dir for section 2
output_dir_autosome_roh_freq <- Sys.getenv("autosome_roh_freq_directory")
output_dir_autosome_roh_freq
output_dir_roh_hotspots_freq <- Sys.getenv("roh_hotspots_freq_directory")

Sys.getenv()

# # Verify the current working directory
# getwd()
```
#1: ROH-Hotspots extraction
```{r ROH-Hotspot extraction}
# Load the MASS package for fitdistr function
library(MASS)
library(ggplot2)
# Read the data without treating "#" as a comment
data <- read.table(input_file, header = TRUE, sep = "\t", comment.char = "")
# Assigning column names
colnames(data) <- c("CHR", "POS1", "POS2", "COUNT", "FREQUENCY")
if (simulation_processing) { 
  # Find the row where max_length is inside the range (POS1, POS2)
  last_mapped_window <- data[data$POS1 <= max_length_chromosome & data$POS2 >= max_length_chromosome, ]

  # Check if a row was found
  if (nrow(last_mapped_window) > 0) {
      pos2_last_mapped_window <- last_mapped_window$POS2[1]
      # Remove all rows where POS2 is larger than pos2_last_mapped_window
      data <- data[data$POS2 <= pos2_last_mapped_window, ]
      
  } else {
      cat("No row found containing max_length.\n")
  }

} 
# Extracting the frequency data for the genomic windows
values <- data$FREQUENCY
# Sort the values in descending order to find the top 1%
sorted_values <- sort(values, decreasing = TRUE)
# Calculate the threshold for the top 1%
threshold_index <- ceiling(roh_hotspot_threshold_percentage * length(sorted_values))
threshold <- sorted_values[threshold_index]
# Identify genomic windows with ROH frequency above the threshold
top_1_percent_ROH <- data[data$FREQUENCY >= threshold, ]
# Print the top 1% of ROH regions
print(top_1_percent_ROH)
```
## 1.1: Gapless ROH-hotspot windows 
Here, directly adjacent ROH-hotspot windows are merged together.
For these ROH-hotspot windows, no gaps are allowed within the ROH-hotspot window.
```{r Adjacent ROH-hotspot windows merge}
# Initialize an empty dataframe to store merged rows
merged_top_1_percent_ROH <- data.frame(CHR = character(), POS1 = integer(), POS2 = integer(), COUNT = integer(), FREQUENCY = numeric())
# Initialize a variable to store the previous row
prev_row <- NULL

# Iterate over each row in the top 1% ROH data
for (i in 1:nrow(top_1_percent_ROH)) {
  # Get the current row
  current_row <- top_1_percent_ROH[i, ]
  
  # Check if it's the first row or if conditions for merging are met
  if (!is.null(prev_row) && current_row$CHR == prev_row$CHR && current_row$COUNT == prev_row$COUNT && as.integer(current_row$POS1) == as.integer(prev_row$POS2) + 1) {
    # Merge the current row with the previous one
    prev_row$POS2 <- current_row$POS2
  } else {
    # If conditions for merging are not met, add the previous row to the merged dataframe
    if (!is.null(prev_row)) {
      merged_top_1_percent_ROH <- rbind(merged_top_1_percent_ROH, prev_row)
    }
    # Update the previous row to the current row
    prev_row <- current_row
  }
}

# Add the last row to the merged dataframe
if (!is.null(prev_row)) {
  merged_top_1_percent_ROH <- rbind(merged_top_1_percent_ROH, prev_row)
}
# Calculate the length in Megabases (Mb)
merged_top_1_percent_ROH$Length_mb <- (merged_top_1_percent_ROH$POS2 - merged_top_1_percent_ROH$POS1 + 1) / 10^6
# Compute distance to previous and next window
merged_top_1_percent_ROH$Distance_to_prev_kb <- c(NA, diff(merged_top_1_percent_ROH$POS1)/1000)
merged_top_1_percent_ROH$Distance_to_next_kb <- c(diff(merged_top_1_percent_ROH$POS1)/1000, NA)

# Print the merged top 1% ROH data
print(merged_top_1_percent_ROH)
```
## 1.2: Merging together ROH-hotspot windows within 2Mb
For these ROH-hotspot windows, gaps of up to 2 Megabases are allowed, when linking together "up to 2Mb adjacent" ROH-hotspot windows.
```{r ROH-hotspot windows within 2Mb merge}
# Initialize an empty dataframe to store linked windows
linked_hotspot_windows <- data.frame(CHR = character(), POS1 = integer(), POS2 = integer(), Length_Mb = numeric())
# Initialize variables to keep track of linked window
current_window <- NULL

# Iterate over each row in the merged dataframe
for (i in 1:nrow(merged_top_1_percent_ROH)) {
  # Get the current row
  current_row <- merged_top_1_percent_ROH[i, ]
  
  # If it's the first row or if it's not within the specified distance from the previous window, start a new window
  if (is.null(current_window) || current_row$POS1 - current_window$POS2 > max_hotspot_windows_gap || current_row$CHR != current_window$CHR) {
    if (!is.null(current_window)) {
      # Calculate Length_kb for the completed window and add it to the linked dataframe
      current_window$Length_Mb <- (current_window$POS2 - current_window$POS1 + 1) / 10^6
      linked_hotspot_windows <- rbind(linked_hotspot_windows, current_window[, c("CHR", "POS1", "POS2", "Length_Mb")])
    }
    # Start a new window
    current_window <- current_row
  } else {
    # Extend the current window
    current_window$POS2 <- current_row$POS2
  }
}


# Add the last window to the linked dataframe
if (!is.null(current_window)) {
  current_window$Length_Mb <- (current_window$POS2 - current_window$POS1 + 1) / 10^6
  linked_hotspot_windows <- rbind(linked_hotspot_windows, current_window[, c("CHR", "POS1", "POS2", "Length_Mb")])
}

# Print the linked windows
print(linked_hotspot_windows)
```
## 1.3 Creating the final ROH-hotspots by adding adjacent windows under the hotspot-threshold
Adjacent 100kbp subwindows with ROH-frequencies >= maximum threshold of a window - 0.1, are incorporated to the hotspots in each direction, until noamore subwindows not reaching this criteria are not found. 
```{r}
# Assuming merged_top_1_percent_ROH and data are data frames
incorporate_adjacent_windows <- function(merged_df, data_df, threshold) {
  # new_threshold <- threshold - 0.1
  merged_df$Incorporated <- FALSE  # Track incorporated windows
  
  for (i in 1:nrow(merged_df)) {
    current_row <- merged_df[i, ]
    chrom <- current_row$CHR
    pos1 <- current_row$POS1
    pos2 <- current_row$POS2
    # max(data$FREQUENCY[data$CHR == 1 & data$POS1 >= 21900001 & data$POS2 <= 24000000])
    threshold_frequency <- max(data_df$FREQUENCY[data_df$CHR == chrom & data_df$POS1 >= pos1 & data_df$POS2 <= pos2])
    new_threshold <- threshold_frequency - 0.1
    # Find corresponding 100kbp window in data
    data_pos1_index <- which(data_df$CHR == chrom & data_df$POS1 == pos1)

    if (length(data_pos1_index) > 0) {
      index <- data_pos1_index[1]
      # Check lower index condition
      while (index - 1 >= 1) {
        lower_row <- data_df[index - 1, ]
        if (lower_row$CHR == chrom && lower_row$FREQUENCY >= new_threshold) {
          # Incorporate this window
          merged_df[i, "POS1"] <- lower_row$POS1
          merged_df[i, "Incorporated"] <- TRUE
          index <- index - 1
        } else {
          break
        }
      }
    }
    
    # Check if the next ROH-hotspot is close enough to skip POS2 check
    if (i < nrow(merged_df)) {
      next_row <- merged_df[i+1, ]
      if (next_row$POS1 - pos2 > 10^5) {
        # Find corresponding 100kbp window for POS2
        data_pos2_index <- which(data_df$CHR == chrom & data_df$POS2 == pos2)
        if (length(data_pos2_index) > 0) {
          index <- data_pos2_index[1]
          # Check higher index condition
          while (index + 1 <= nrow(data_df)) {
            higher_row <- data_df[index + 1, ]
            if (higher_row$CHR == chrom && higher_row$FREQUENCY >= new_threshold) {
              # Incorporate this window
              merged_df[i, "POS2"] <- higher_row$POS2
              merged_df[i, "Incorporated"] <- TRUE
              index <- index + 1
            } else {
              break
            }
          }
        }
      }
    }
  }
  return(merged_df)
}
ROH_hotspots_df <- incorporate_adjacent_windows(linked_hotspot_windows, data, threshold)
ROH_hotspots_df$Length_Mb <- (ROH_hotspots_df$POS2-ROH_hotspots_df$POS1+1)/10^6
print(ROH_hotspots_df)
```
## 1.4 Extracting the ROH-frequencies for the final ROH-hotspot windows
```{r}
# Initialize an empty dataframe to store linked windows
ROH_hotspot_windows_frequencies <- data.frame(CHR = character(), POS1 = integer(), POS2 = integer(),COUNT = integer(),FREQUENCY = numeric())
for (i in 1:nrow(ROH_hotspots_df)) {
   ROH_hotspot_100kbp_windows <- data[data$CHR == ROH_hotspots_df$CHR[i] & data$POS1 >= ROH_hotspots_df$POS1[i] & data$POS2 <= ROH_hotspots_df$POS2[i], ]
   ROH_hotspot_100kbp_windows$Hotspot_interval <- c(rep(paste0("[", ROH_hotspots_df$POS1[i], ",", ROH_hotspots_df$POS2[i], "]"),nrow(ROH_hotspot_100kbp_windows)))
   ROH_hotspot_100kbp_windows$Length_Mb <- (ROH_hotspots_df$POS2[i] - ROH_hotspots_df$POS1[i] + 1) / 10^6
    
   ROH_hotspot_windows_frequencies <- rbind(ROH_hotspot_windows_frequencies,ROH_hotspot_100kbp_windows)
 }
ROH_hotspot_windows_frequencies
```
## 1.5: Visualization of ROH Hotspots
```{r ROH hotspots visualization}
# # Create scatter plots for each chromosome
# for (CHR in unique(data$CHR)) {
#   # Subset data for the current chromosome
#   chr_data <- data[data$CHR == CHR, ]
#   chr_merged_data <- ROH_hotspot_windows_frequencies[ROH_hotspot_windows_frequencies$CHR == CHR, ]
#   
#   # Create scatter plot
#   p <- ggplot() +
#     geom_point(data = chr_data, aes(x = POS1, y = FREQUENCY, color = "ROH-segments"), alpha = 0.5) +
#     geom_point(data = chr_merged_data, aes(x = POS1, y = FREQUENCY, color = "ROH hotspots"), alpha = 0.5) +
#     geom_hline(aes(yintercept = threshold, color = "ROH hotspot threshold"), linetype = "dashed") +
#     labs(title = paste("Scatter Plot for Chromosome", CHR),
#          x = "Position",
#          y = "FREQUENCY",
#          color = "Legend") +
#     scale_color_manual(values = c("ROH-segments" = "gray", "ROH hotspots" = "blue", "ROH hotspot threshold" = "red"),
#                        labels = c("ROH-segments" = "ROH-segments", "ROH hotspots" = "ROH hotspots", "ROH hotspot threshold" = "ROH hotspot threshold")) +
#     theme_minimal()
# 
#   # Print the plot
#   print(p)
# }
```
# 2: Exporting ROH-frequency files for Gosling Visualizations
## 2.0: Formatting the Threshold for the filename
```{r ,eval=TRUE}
# formatted_threshold <- format(threshold, nsmall = 3)
formatted_threshold <- round(threshold,5)
```
## 2.1: ROH Hotspot-frequency into .tsv files
```{r Export ROH Hotspot-frequency,eval=TRUE}
# Define a function to split each window into two rows
split_window <- function(data) {
  # Create a new dataframe to store the split rows
  new_data <- data.frame(
    CHR = rep(data$CHR,2),
    POS = c(data$POS1, data$POS2),
    COUNT = rep(data$COUNT, 2),
    FREQUENCY = rep(data$FREQUENCY, 2),
    Length_Mb = rep(data$Length_Mb, 2),
    Hotspot_interval = rep(data$Hotspot_interval,2)
  )
  return(new_data)
}

# Initialize an empty list to store the split dataframes
split_data <- list()
# Iterate over each chromosome
for (Chromosome in unique(ROH_hotspot_windows_frequencies$CHR)) {
  # Subset data for the current chromosome
  chr_data <- ROH_hotspot_windows_frequencies[ROH_hotspot_windows_frequencies$CHR == Chromosome, ]
  # Split each window into two rows
  split_data[[Chromosome]] <- split_window(chr_data)
}
# Combine all split dataframes into one
split_data_combined <- do.call(rbind, split_data)

# Export the combined split dataframe to TSV files for each chromosome
for (Chromosome in unique(ROH_hotspot_windows_frequencies$CHR)) {
  # Subset the split data for the current chromosome
  chr_split_data <- split_data_combined[split_data_combined$CHR == Chromosome, ]
  
  # Checking if the prefix of the input file contains "chr" And Naming the outputfile 
  if (grepl("_chr[0-9]+", prefix)) {
    # If '_chr' followed by a number exists in the prefix
    filename <- file.path(output_dir_roh_hotspots_freq, paste(prefix, "_ROH_Hotspot_freq.tsv", sep = ""))
  } else {
    # If '_chr' followed by a number does not exist in the prefix
    filename <- file.path(output_dir_roh_hotspots_freq, paste(prefix, "_chr", Chromosome, "_ROH_Hotspot_freq.tsv", sep = ""))
  }
  
  # Create the directory if it doesn't exist
  dir.create(dirname(filename), recursive = TRUE, showWarnings = FALSE)
    
  # Write data to TSV file without quotes and with tab separation
  write.table(chr_split_data, file = filename, sep = "\t", row.names = FALSE, quote = FALSE)
  
  # Print message indicating successful export
  cat("Exported ROH-hotspot ROH-frequencies to", filename, "\n")
}
```
## 2.2: ROH-frequency into .tsv files (For all chr)
```{r Export ROH-frequency for all chromosomes}
# Define a function to split each window into two rows
split_window <- function(data) {
  # Create a new dataframe to store the split rows
  new_data <- data.frame(
    CHR = data$CHR,
    POS = c(data$POS1, data$POS2),
    COUNT = rep(data$COUNT, 2),
    FREQUENCY = rep(data$FREQUENCY, 2)
  )
  return(new_data)
}

# Initialize an empty list to store the split dataframes
split_data <- list()

# Iterate over each chromosome
for (Chromosome in unique(data$CHR)) {
  # Subset data for the current chromosome
  chr_data <- data[data$CHR == Chromosome, ]
  
  # Split each window into two rows
  split_data[[Chromosome]] <- split_window(chr_data)
}

# Combine all split dataframes into one
split_data_combined <- do.call(rbind, split_data)

# Export the combined split dataframe to TSV files for each chromosome
for (Chromosome in unique(data$CHR)) {
  # Subset the split data for the current chromosome
  chr_split_data <- split_data_combined[split_data_combined$CHR == Chromosome, ]   
  
  # Checking if the prefix of the input file contains "chr" And Naming the outputfile 
  if (grepl("_chr[0-9]+", prefix)) {
    # If '_chr' followed by a number exists in the prefix
    filename <- file.path(output_dir_autosome_roh_freq, paste(prefix,"_ROH_freq", "_threshold_", formatted_threshold,".tsv", sep = ""))
  } else {
    # If '_chr' followed by a number does not exist in the prefix
    filename <- file.path(output_dir_autosome_roh_freq, paste(prefix,"_chr", Chromosome,"_ROH_freq", "_threshold_", formatted_threshold,".tsv", sep = ""))
  }
  # Write data to TSV file without quotes and with tab separation
  write.table(chr_split_data, file = filename, sep = "\t", row.names = FALSE, quote = FALSE)
  
  # Print message indicating successful export
  cat("Exported autosome ROH-frequencies to", filename, "\n")
}
```
# 3 Exporting ROH-hotspot windows into .bed-files
## 3.2: Gapless ROH-hotspot windows 
```{r Export Gapless ROH-hotpots to .bed-files}
# Iterate over unique chromosomes
for (Chromosome in unique(ROH_hotspot_windows_frequencies$CHR)) {
  # Subset data for the current chromosome
  chr_data <- ROH_hotspot_windows_frequencies[ROH_hotspot_windows_frequencies$CHR == Chromosome, c("CHR", "POS1", "POS2")]
  
  # Checking if the prefix of the input file contains "chr" And Naming the outputfile 
  if (grepl("_chr[0-9]+", prefix)) {
    # If '_chr' followed by a number exists in the prefix
    filename <- file.path(output_dir_gapless_roh_hotspots, paste(prefix,"_ROH_Hotspot_gapless_windows.bed", sep = ""))
  } else {
    # If '_chr' followed by a number does not exist in the prefix
    # Define the filename with the output directory path
    filename <- file.path(output_dir_gapless_roh_hotspots, paste(prefix,"_chr", Chromosome,"_ROH_Hotspot_gapless_windows.bed", sep = "")) 
  }
  # Write data to .bed file
  write.table(chr_data, file = filename, sep = "\t", quote = FALSE, col.names = FALSE, row.names = FALSE)
  
  # Print message indicating successful export
  cat("Exported", nrow(chr_data), "hotspot windows to", filename, "\n")
}
```
## 3.2: 2Mb ROH-hotspot windows
```{r Export ROH-hotpots to .bed-files}
# Iterate over unique chromosomes
for (Chromosome in unique(ROH_hotspots_df$CHR)) {
  # Subset data for the current chromosome
  chr_data <- ROH_hotspots_df[ROH_hotspots_df$CHR == Chromosome, c("CHR", "POS1", "POS2")]
  
  # Checking if the prefix of the input file contains "chr" And Naming the outputfile 
  if (grepl("_chr[0-9]+", prefix)) {
    # If '_chr' followed by a number exists in the prefix
    filename <- file.path(output_dir, paste(prefix,"_ROH_Hotspot_windows.bed", sep = "")) 
  } else {
    # If '_chr' followed by a number does not exist in the prefix
    # Define the filename with the output directory path
    filename <- file.path(output_dir, paste(prefix,"_chr", Chromosome,"_ROH_Hotspot_windows.bed", sep = "")) 
  }
  # Write data to .bed file
  write.table(chr_data, file = filename, sep = "\t", quote = FALSE, col.names = FALSE, row.names = FALSE)
  
  # Print message indicating successful export
  cat("Exported", nrow(chr_data), "hotspot windows to", filename, "\n")
}
```
