---
title: "Founder population simulation for Dogs (Neutral Model) in AlphaSimR"
output:
  html_document:
    toc: true
    toc_depth: 3  
date: "2024-04-19"
editor_options: 
  markdown: 
    wrap: 72
---

# 0: Preparation

## Defining the output directory & the chromosome to be simulated

```{r}
# Clean the working environment
rm(list = ls())

knitr::opts_chunk$set(echo = TRUE)



####################################  
# Defining the working directory
#################################### 

# Set the path to your GitHub folder
YOUR_GITHUB_ROOT_DIRECTORY <- "C:/Users/jonat/GitHub"

# Defining the relative path in the repository
repository_path <- "Computational-modelling-of-genomic-inbreeding-and-roh-islands-in-extremely-small-populations"

output_dir_simulation <- file.path(YOUR_GITHUB_ROOT_DIRECTORY, repository_path,"data/raw/simulated")


if (!dir.exists(output_dir_simulation)) {
  # Create the working directory if it doesn't exist
  dir.create(output_dir_simulation)
}

# Set the working directory for notebook chunks
knitr::opts_knit$set(root.dir = output_dir_simulation)

#################################### 
# Defining Input parameters
#################################### 
min_MAF <- 0.05 #(minimum allowed Minor Allele Frequency for each SNP)



#################################### 
# Defining the output files
#################################### 


# Define the chromosome to be simulated
chr_simulated <- "chr3"

# Define the base name for the output .map & .ped PLINK files
output_sim_files_basename <- paste0("Neutral_simulation_", chr_simulated)


# # Verify the current working directory
#getwd()
```

## Loading libraries

```{r library()}
library(AlphaSimR)
library(knitr)


#library(dplyr)
#library(ggplot2)
#library(patchwork)
#library(purrr)
#library(tibble)

```

# 1: Creating founder Haplotypes

## 1.0: Defining chromosome lengths of model species

```{r}
# Chromosome lengths of Dogs in Mb, derived from Table 1 in this article:
# https://www-ncbi-nlm-nih-gov.ezproxy.its.uu.se/pmc/articles/PMC2564286/

chromosome_lengths_mb <- c(
    "chr1" = 125, "chr2" = 88, "chr3" = 94, "chr4" = 91, "chr5" = 91,
    "chr6" = 80, "chr7" = 83, "chr8" = 77, "chr9" = 64, "chr10" = 72,
    "chr11" = 77, "chr12" = 75, "chr13" = 66, "chr14" = 63, "chr15" = 67,
    "chr16" = 62, "chr17" = 67, "chr18" = 58, "chr19" = 56, "chr20" = 61,
    "chr21" = 54, "chr22" = 64, "chr23" = 55, "chr24" = 50, "chr25" = 54,
    "chr26" = 42, "chr27" = 48, "chr28" = 44, "chr29" = 44, "chr30" = 43,
    "chr31" = 42, "chr32" = 41, "chr33" = 34, "chr34" = 45, "chr35" = 29,
    "chr36" = 33, "chr37" = 33, "chr38" = 26
)
```

## 1.1: runMacs2() - Generating founder population

`runMacs2` Parameters:

-   nInd = 100: 100 individuals in the founder population

-   nChr = 1: Haplotypes are created for 1 chromosome per individual

-   Ne = 2500: Effective population size: 2500 (Ne)

-   bp: base pair length of chromosome

-   genLen = 1: Genetic length of chromosome in Morgans, set as 1.

-   HistNe = NULL: No effective population size defined from previous
    generations (histNe = NULL)

-   histGen = NULL: If HistNe was defined, this parameter would define
    the number of generations ago we had the Ne defined in HistNe

```{r runMacs2()}
help(runMacs2)

model_chromosome_bp_length <- chromosome_lengths_mb[chr_simulated] * 10^6

founderGenomes <- runMacs2(nInd = 100,
                     nChr = 1,
                     Ne = 2500,
                     bp = model_chromosome_bp_length,
                     genLen = 1,
                     histNe = NULL,
                     histGen = NULL)

# Inspecting the founderGenomes object
founderGenomes

```

## 1.2: SimParam() - Setting Global Simulation Parameters

```{r SimParam()}

help(SimParam)

# Creating a new Simparam object & Assigning the founder population population to it
SP <- SimParam$new(founderGenomes) 
# SP

```



### 1.2.1: \$setTrackRec() - Setting on recombination tracking for the simulation

Sets recombination tracking for the simulation. By default recombination
tracking is turned off. **When turned on recombination tracking will
also turn on pedigree tracking.**

**Recombination tracking keeps records of all individuals created,**
except those created by hybridCross, because their pedigree is not
tracked.

```{r SimParam$setTrackRec()}
SP$setTrackRec(TRUE)
```

## 1.3: newPop() - Creating two separate populations for bottleneck event

```{r newPop()}
# Generate the initial founder population
founderpop <- newPop(founderGenomes, simParam = SP)
# Set misc used for setting years of birth for the first individuals
current_generation = 0
founderpop = setMisc(x = founderpop,
                   node = "generation",
                   value = current_generation)
head(getMisc(x = founderpop, node = "generation"))


# Split the founder population into two breeding groups
breed1_founders <- founderpop[1:50]
breed1_founders

breed2_founders <- founderpop[51:100]
breed2_founders

```

# 2: Forward in time simulation

## 2.1: randCross() - Bottleneck Simulation for 10 Generations

`randCross()` simulates random mating.

-   **nCrosses**-argument specifies how many times an individual in the
    population can be a parent. In this case, an individual can only be
    a parent once (random mating)

-   **nProgeny**-argument specifies how many progeny's each mating pair
    can have. In this case, this parameter is set at 1 progeny per
    mating pair.

-   **simParam**-argument specifies the global simulation parameters

```{r}
random_mating <- function(pop, SP, n_gen) {
    n_ind <- pop@nInd # Extracting number of individuals in the current population
    generations <- vector(length = n_gen + 1,
                          mode = "list") 
    generations[[1]] <- pop # The initial breeding group population gets stored as the first element
    
    # Simulating randomm mating to perform simulation of the 2nd until the n+1:th generation
    # Each mating Progenys are generated using randcross
    for (gen_ix in 2:(n_gen + 1)) {
        generations[[gen_ix]] <- randCross(pop=generations[[gen_ix - 1]], nCrosses = n_ind, nProgeny = 1,
                                           simParam = SP)
    }
     bottleneck_generations <- generations[-1] #all generations except for the founder population gets returned (the simulated generations derived from the bottleneck)
    
    return(bottleneck_generations)
}

n_generations <- 5
# Simulating random mating within each breeding group for 5 generations
breed1_bottleneck <- random_mating(breed1_founders, SP, n_generations)
breed2_bottleneck <- random_mating(breed2_founders, SP, n_generations)
```

Viewing the output from the bottleneck simulation

```{r}

# founders
founderpop
breed1_bottleneck
```

## 2.2 Extracting the final generation

```{r}
# Extracting final generation:
last_generation <- breed1_bottleneck[[n_generations]]
last_generation
```
## 2.3: Adding SNPs to the final generation ( SimParam$addSnpChip())

### 2.3.1 Single chromosome usage: Computing number of SNPs to assign to a chromosome

```{r}

# Sum up the chromosome lengths to get the autosome length
autosome_length <- sum(chromosome_lengths_mb)

# Number of SNP markers from the empirical dataset
# https://datadryad.org/stash/dataset/doi:10.5061/dryad.h44j0zpkf
num_snp_markers <- 78088  

# Calculate average SNP-autosome density
average_snp_autosome_density <- num_snp_markers / autosome_length

# Calculate the number of SNP markers to choose for the specific chromosome
num_snp_markers_chr3 <- average_snp_autosome_density * chromosome_lengths_mb[chr_simulated]



# Output results
cat("autosome length:", autosome_length, "Mb\n")
cat("Average SNP-autosome density:", average_snp_autosome_density, "SNPs/Mb\n")
cat("Number of SNP markers for", chr_simulated, ":", num_snp_markers_chr3, "SNPs\n")

```

```{r}
average_chr_SNP_density <- num_snp_markers / length(chromosome_lengths_mb) 
cat(" Average Chr SNP density:", average_chr_SNP_density, "SNPs/Chr\n")

```

### 2.3.2: SimParam$addSnpChip() - Randomly assigns eligible SNPs to a SNP chip

`SP$addSnpChip()`Randomly assigns eligible SNPs to a SNP chip.

**nSnpPerChr**-argument that assigns for each individuals n SNPs for
each chromsome, where n is the user-specified number. 

**minSnpFreq**-argument sets the Minimum Allele Frequency (MAF) allowed for each marker.
For instance, setting minSnpFreq = 0.05 ensures that all markers will have MAF >= 0.05 

**refPop**-argument specifies which population will be genotyped.
In this case, it will be the last_generation after the simulated population bottleneck scenario.

```{r SimParam$addSnpChip()}
SP$addSnpChip(
  nSnpPerChr = num_snp_markers_chr3,
  minSnpFreq = min_MAF,
  refPop = last_generation
  ) 
```


# 3: Export to PLINK Extracting data to PLINK

## 3.0.1: Optional: pullSnpGeno() - Retrieves a list of the existing marker id:s (SNP id:s)

Function that retrieves SNP genotype data & returns a matrix of SNP
genotypes

```{r pullSnpGeno(), echo = FALSE}
#pullSnpGeno(last_generation,simParam=SP)
```

## 3.0.2: Optional: getSnPMap() - Retrieves a table of the genetic map

Function that retrieves the genetic map for a given SNP chip & returns a
data.frame with:

-   **id:** Unique identifier for the marker (SNP)

-   **chr:** Chromosome containing the SNP

-   **site:** Segregating site on the chromosome

-   **pos:** Genetic map position

```{r getSnPMap()}
# SP$addSnpChip(5)
# getSnpMap(snpChip=1,simParam=SP) 

#getSnpMap(simParam=SP)

cat("First 10 markers and their genetic map positions:\n")
kable(head(getSnpMap(simParam = SP), 10))
#head(getSnpMap(simParam = SP), 10)


```

## 3.1: writePlink()

**Description** Function that inputs a Pop-class & exports it to PLINK
PED and MAP files.

The arguments for this function were chosen for consistency with
RRBLUP2. **The base pair coordinate will the locus position as stored in
AlphaSimR** & not an actual base pair position, because AlphaSimR
doesn’t track base pair positions, **only relative positions for the
loci used in the simulation**.

\*\* Usage \*\* writePlink( pop, baseName, traits = 1, use = "pheno",
snpChip = 1, useQtl = FALSE, simParam = NULL, ... )

**Arguments:**

-   **pop** an object of Pop-class

-   **baseName** basename for PED and MAP files.

-   **traits** an integer indicating the trait to write, a trait name,
    or a function of the traits returning a single value.

-   **use** what to use for PLINK’s phenotype field. Either phenotypes
    "pheno", genetic values "gv", estimated breeding values "ebv",
    breeding values "bv", or random values "rand".

-   **snpChip** an integer indicating which SNP chip genotype to use
    useQtl should QTL genotypes be used instead of a SNP chip.

    ```         
       -    If TRUE, snpChip specifies which trait’s QTL to use, and thus these QTL may not match the QTL underlying the phenotype supplied in traits.
    ```

-   **simParam** an object of SimParam ... additional arguments if using
    a function for traits

```{r}
# Setting the directory where the .map and .ped files will be stored in
setwd(output_dir_simulation)

writePlink(last_generation,simParam=SP, baseName=output_sim_files_basename,
           
            traits = 1,
           use = "rand"
           
           
           )
```

# 4: Convert distances to basepair in the .map file

```{r}
# Setting the working directory as the output directory
setwd(output_dir_simulation)

# Read PLINK files into R objects
ped_file <- paste0(output_sim_files_basename, '.ped')
map_file <- paste0(output_sim_files_basename, '.map')
ped_data <- read.table(ped_file)
map_data <- read.table(map_file)


genetic_distance_morgan_column <- map_data$V3 # Extracting the genetic distance column of map_data (3rd column) 
base_pair_position_column <- round(genetic_distance_morgan_column * (model_chromosome_bp_length / 100))

# Redefining the physical position column (4th column) with the new values defined in base_pair_position_column
map_data$V4 <- base_pair_position_column

write.table(map_data, file = map_file, sep = "\t", row.names = FALSE, col.names = FALSE,quote = FALSE)

#col.names = FALSE: Removes the header.
#quote = FALSE: Removes quotation marks from values.


cat("Showing the first 10 rows of the new .map file:\n")
kable(head(map_data, 10))

```
