---
title: "ROH-hotspot identification"
output:
  html_document:
    toc: true
    toc_depth: 3  
    output_dir: output_dir
date: "2024-05-07"
editor_options: 
  markdown: 
    wrap: 72
---

# 0: Preparation
Defining the input and output files

```{r setup}
# Clean the working environment
rm(list = ls())

knitr::opts_chunk$set(echo = TRUE)

####################################  
# Defining the working directory
#################################### 

# Set the path to your GitHub folder
YOUR_GITHUB_ROOT_DIRECTORY <- "C:/Users/jonat/GitHub"

# Defining the relative path in the repository
repository_path <- file.path(YOUR_GITHUB_ROOT_DIRECTORY,"Computational-modelling-of-genomic-inbreeding-and-roh-islands-in-extremely-small-populations")

results_dir <- file.path(repository_path,"res_test")
empirical_dog_breed <- "labrador_retriever"

####################################  
# Defining the input file
#################################### 

Coverage_dir <- file.path(results_dir,"Bedtools/coverage")
pop_roh_freq_dir <- file.path(Coverage_dir,"empirical",empirical_dog_breed,"pop_roh_freq")

# input_file <- file.path(repository_path,"res_test/Bedtools/coverage/empirical/doi_10_5061_dryad_h44j0zpkf__v20210813/pop_roh_freq/german_shepherd_ROH_freq.bed")
input_file <- file.path(pop_roh_freq_dir,"labrador_retriever_ROH_freq.bed")

## Extracting the prefix of the file used  
# Extracting the basename (filename without extension) from the file path
filename <- basename(input_file)
# Split the filename by "_ROH_freq.bed" and get the first part
prefix <- strsplit(filename, "_ROH_freq.bed")[[1]][1]
# Output the prefix
print(prefix)



#################################### 
# Defining the output dirs
#################################### 

output_dir <- file.path(results_dir,"ROH-Hotspots","empirical",empirical_dog_breed)
# Set the working directory for notebook chunks
knitr::opts_knit$set(root.dir = output_dir)

# Output dir for section 3.2
output_dir_gapless_roh_hotspots <- file.path(output_dir, "gapless_roh_hotspots")
if (!dir.exists(output_dir_gapless_roh_hotspots)) {
  # Create the working directory if it doesn't exist
  dir.create(output_dir_gapless_roh_hotspots, recursive = TRUE)
}

# Output dir for section 2
output_dir_gosling_plots <- file.path(output_dir, "Gosling_plots")

output_dir_autosome_roh_freq <- file.path(output_dir_gosling_plots, "autosome_roh_freq")


if (!dir.exists(output_dir_autosome_roh_freq)) {
  # Create the working directory if it doesn't exist
  dir.create(output_dir_autosome_roh_freq, recursive = TRUE)
}
print(output_dir_autosome_roh_freq)

output_dir_roh_hotspots_freq <- file.path(output_dir_gosling_plots, "roh_hotspots_freq")
if (!dir.exists(output_dir_roh_hotspots_freq)) {
  # Create the working directory if it doesn't exist
  dir.create(output_dir_roh_hotspots_freq, recursive = TRUE)
}

# Set the working directory for notebook chunks
knitr::opts_knit$set(root.dir = output_dir)

#################################### 
# Defining Input parameters
#################################### 
roh_hotspot_threshold_percentage <- 0.01 # 0.01 represents top 1 % most homozygous region
max_hotspot_windows_gap <- 2*10^6 # ROH-hotspot windows that are less than 2Mb apart are merged together (see section 1.3)

# # Verify the current working directory
# getwd()

```

#1: ROH-Hotspots extraction

```{r ROH-Hotspot extraction}
# Load the MASS package for fitdistr function
library(MASS)
library(ggplot2)

# Read the data without treating "#" as a comment
data <- read.table(input_file, header = TRUE, sep = "\t", comment.char = "")

# Assigning column names
colnames(data) <- c("CHR", "POS1", "POS2", "COUNT", "FREQUENCY")


# Frequency Data is chosen
values <- data$FREQUENCY

# Sort the values in descending order to find the top 1%
sorted_values <- sort(values, decreasing = TRUE)

# Calculate the threshold for the top 1%
threshold_index <- ceiling(roh_hotspot_threshold_percentage * length(sorted_values))
threshold <- sorted_values[threshold_index]

# Identify genomic windows with ROH frequency above the threshold
top_1_percent_ROH <- data[data$FREQUENCY >= threshold, ]

# Print the top 1% of ROH regions
print(top_1_percent_ROH)
```

## 1.2: Gapless ROH-hotspot windows 
Here, directly adjacent ROH-hotspot windows are merged together.
For these ROH-hotspot windows, no gaps are allowed within the ROH-hotspot window.

```{r Adjacent ROH-hotspot windows merge}
# Initialize an empty dataframe to store merged rows
merged_top_1_percent_ROH <- data.frame(CHR = character(), POS1 = integer(), POS2 = integer(), COUNT = integer(), FREQUENCY = numeric())
# Initialize a variable to store the previous row
prev_row <- NULL

# Iterate over each row in the top 1% ROH data
for (i in 1:nrow(top_1_percent_ROH)) {
  # Get the current row
  current_row <- top_1_percent_ROH[i, ]
  
  # Check if it's the first row or if conditions for merging are met
  if (!is.null(prev_row) && current_row$CHR == prev_row$CHR && current_row$COUNT == prev_row$COUNT && as.integer(current_row$POS1) == as.integer(prev_row$POS2) + 1) {
    # Merge the current row with the previous one
    prev_row$POS2 <- current_row$POS2
  } else {
    # If conditions for merging are not met, add the previous row to the merged dataframe
    if (!is.null(prev_row)) {
      merged_top_1_percent_ROH <- rbind(merged_top_1_percent_ROH, prev_row)
    }
    # Update the previous row to the current row
    prev_row <- current_row
  }
}

# Add the last row to the merged dataframe
if (!is.null(prev_row)) {
  merged_top_1_percent_ROH <- rbind(merged_top_1_percent_ROH, prev_row)
}

# Calculate the length in kilobases (kb)
merged_top_1_percent_ROH$Length_kb <- (merged_top_1_percent_ROH$POS2 - merged_top_1_percent_ROH$POS1 + 1) / 1000

# Compute distance to previous and next window
merged_top_1_percent_ROH$Distance_to_prev_kb <- c(NA, diff(merged_top_1_percent_ROH$POS1)/1000)
merged_top_1_percent_ROH$Distance_to_next_kb <- c(diff(merged_top_1_percent_ROH$POS1)/1000, NA)

# Print the merged top 1% ROH data
print(merged_top_1_percent_ROH)

```

## 1.3 Expanding hotspots thats 10 % lower!
```{r}
# Assuming merged_top_1_percent_ROH and data are data frames

incorporate_adjacent_windows <- function(merged_df, data_df, threshold) {
  merged_df$Incorporated <- FALSE  # Track incorporated windows
  
  for (i in 1:nrow(merged_df)) {
    current_row <- merged_df[i, ]
    chrom <- current_row$CHR
    pos1 <- current_row$POS1
    pos2 <- current_row$POS2

    # Find corresponding 100kbp window in data
    data_pos1_index <- which(data_df$CHR == chrom & data_df$POS1 == pos1)

    if (length(data_pos1_index) > 0) {
      index <- data_pos1_index[1]
      
      # Check lower index condition
      while (index - 1 >= 1) {
        lower_row <- data_df[index - 1, ]
        if (lower_row$FREQUENCY >= threshold - 10) {
          # Incorporate this window
          merged_df[i, "POS1"] <- lower_row$POS1
          merged_df[i, "Incorporated"] <- TRUE
          index <- index - 1
        } else {
          break
        }
      }
    }
    
    # Check if the next ROH-hotspot is close enough to skip POS2 check
    if (i < nrow(merged_df)) {
      next_row <- merged_df[i + 1, ]
      if (next_row$POS1 - pos2 > 100) {
        # Find corresponding 100kbp window for POS2
        data_pos2_index <- which(data_df$CHR == chrom & data_df$POS1 == pos2)
        
        if (length(data_pos2_index) > 0) {
          index <- data_pos2_index[1]
          
          # Check higher index condition
          while (index + 1 <= nrow(data_df)) {
            higher_row <- data_df[index + 1, ]
            if (higher_row$FREQUENCY >= threshold - 10) {
              # Incorporate this window
              merged_df[i, "POS2"] <- higher_row$POS2
              merged_df[i, "Incorporated"] <- TRUE
              index <- index + 1
            } else {
              break
            }
          }
        }
      }
    }
  }
  
  return(merged_df)
}


result <- incorporate_adjacent_windows(merged_top_1_percent_ROH, data, threshold)
print(result)

```




## 1.4: Merging together ROH-hotspot windows within 2Mb
For these ROH-hotspot windows, gaps of up to 2 Megabases are allowed, when linking together "up to 2Mb adjacent" ROH-hotspot windows.

```{r ROH-hotspot windows within 2Mb merge}

# Initialize an empty dataframe to store linked windows
linked_hotspot_windows <- data.frame(CHR = character(), POS1 = integer(), POS2 = integer(), Length_kb = numeric())

# Initialize variables to keep track of linked window
current_window <- NULL

# Iterate over each row in the merged dataframe
for (i in 1:nrow(merged_top_1_percent_ROH)) {
  # Get the current row
  current_row <- merged_top_1_percent_ROH[i, ]
  
  # If it's the first row or if it's not within the specified distance from the previous window, start a new window
  if (is.null(current_window) || current_row$POS1 - current_window$POS2 > max_hotspot_windows_gap || current_row$CHR != current_window$CHR) {
    if (!is.null(current_window)) {
      # Calculate Length_kb for the completed window and add it to the linked dataframe
      current_window$Length_kb <- (current_window$POS2 - current_window$POS1 + 1) / 1000
      linked_hotspot_windows <- rbind(linked_hotspot_windows, current_window[, c("CHR", "POS1", "POS2", "Length_kb")])
    }
    # Start a new window
    current_window <- current_row
  } else {
    # Extend the current window
    current_window$POS2 <- current_row$POS2
  }
}

# Add the last window to the linked dataframe
if (!is.null(current_window)) {
  current_window$Length_kb <- (current_window$POS2 - current_window$POS1 + 1) / 1000
  linked_hotspot_windows <- rbind(linked_hotspot_windows, current_window[, c("CHR", "POS1", "POS2", "Length_kb")])
}

# Print the linked windows
print(linked_hotspot_windows)



```

## 1.5: Visualization of ROH Hotspots

```{r ROH hotspots visualization}
# Create scatter plots for each chromosome
for (CHR in unique(data$CHR)) {
  # Subset data for the current chromosome
  chr_data <- data[data$CHR == CHR, ]
  chr_merged_data <- merged_top_1_percent_ROH[merged_top_1_percent_ROH$CHR == CHR, ]
  
  # Create scatter plot
  p <- ggplot() +
    geom_point(data = chr_data, aes(x = POS1, y = FREQUENCY, color = "ROH-segments"), alpha = 0.5) +
    geom_point(data = chr_merged_data, aes(x = POS1, y = FREQUENCY, color = "ROH hotspots"), alpha = 0.5) +
    geom_hline(aes(yintercept = threshold, color = "ROH hotspot threshold"), linetype = "dashed") +
    labs(title = paste("Scatter Plot for Chromosome", CHR),
         x = "Position",
         y = "FREQUENCY",
         color = "Legend") +
    scale_color_manual(values = c("ROH-segments" = "gray", "ROH hotspots" = "blue", "ROH hotspot threshold" = "red"),
                       labels = c("ROH-segments" = "ROH-segments", "ROH hotspots" = "ROH hotspots", "ROH hotspot threshold" = "ROH hotspot threshold")) +
    theme_minimal()

  # Print the plot
  print(p)
}


```

# 2: Exporting ROH-frequency files for Gosling Visualizations

## 2.0: Formatting the Threshold for the filename, to three decimals

```{r ,eval=TRUE}
formatted_threshold <- format(threshold, nsmall = 3)
```

## 2.1: ROH Hotspot-frequency into .tsv files

```{r Export ROH Hotspot-frequency,eval=TRUE}
# Define a function to split each window into two rows
split_window <- function(data) {
  # Create a new dataframe to store the split rows
  new_data <- data.frame(
    CHR = data$CHR,
    POS = c(data$POS1, data$POS2),
    COUNT = rep(data$COUNT, 2),
    FREQUENCY = rep(data$FREQUENCY, 2),
    Length_kb = rep(data$Length_kb, 2),
    Hotspot_interval = c(paste0("[", data$POS1, ",", data$POS2, "]"), paste0("[", data$POS1, ",", data$POS2, "]"))
  )
  return(new_data)
}

# Initialize an empty list to store the split dataframes
split_data <- list()

# Iterate over each chromosome
for (Chromosome in unique(merged_top_1_percent_ROH$CHR)) {
  # Subset data for the current chromosome
  chr_data <- merged_top_1_percent_ROH[merged_top_1_percent_ROH$CHR == Chromosome, ]
  
  # Split each window into two rows
  split_data[[Chromosome]] <- split_window(chr_data)
}

# Combine all split dataframes into one
split_data_combined <- do.call(rbind, split_data)

# Export the combined split dataframe to TSV files for each chromosome
for (Chromosome in unique(merged_top_1_percent_ROH$CHR)) {
  # Subset the split data for the current chromosome
  chr_split_data <- split_data_combined[split_data_combined$CHR == Chromosome, ]
  
  # Define the filename with the output directory path
  filename <- file.path(output_dir_roh_hotspots_freq, paste(prefix,"_chr", Chromosome, "_ROH_Hotspot_freq.tsv", sep = ""))
  # Create the directory if it doesn't exist
  dir.create(dirname(filename), recursive = TRUE, showWarnings = FALSE)
    
  # Write data to TSV file without quotes and with tab separation
  write.table(chr_split_data, file = filename, sep = "\t", row.names = FALSE, quote = FALSE)
  

}




```

## 2.2: ROH-frequency into .tsv files (For all chr)

```{r Export ROH-frequency for all chromosomes}
# Define a function to split each window into two rows
split_window <- function(data) {
  # Create a new dataframe to store the split rows
  new_data <- data.frame(
    CHR = data$CHR,
    POS = c(data$POS1, data$POS2),
    COUNT = rep(data$COUNT, 2),
    FREQUENCY = rep(data$FREQUENCY, 2)
  )
  return(new_data)
}

# Initialize an empty list to store the split dataframes
split_data <- list()

# Iterate over each chromosome
for (Chromosome in unique(data$CHR)) {
  # Subset data for the current chromosome
  chr_data <- data[data$CHR == Chromosome, ]
  
  # Split each window into two rows
  split_data[[Chromosome]] <- split_window(chr_data)
}

# Combine all split dataframes into one
split_data_combined <- do.call(rbind, split_data)

# Export the combined split dataframe to TSV files for each chromosome
for (Chromosome in unique(data$CHR)) {
  # Subset the split data for the current chromosome
  chr_split_data <- split_data_combined[split_data_combined$CHR == Chromosome, ]   
  
    # Define the filename with the output directory path
  filename <- file.path(output_dir_autosome_roh_freq, paste(prefix,"_chr", Chromosome,"_ROH_freq", "_threshold_", formatted_threshold,".tsv", sep = ""))
  
  # Write data to TSV file without quotes and with tab separation
  write.table(chr_split_data, file = filename, sep = "\t", row.names = FALSE, quote = FALSE)
}


```

# 3 Exporting ROH-hotspot windows into .bed-files

## 3.2: Gapless ROH-hotspot windows 

```{r Export ROH-hotpots to .bed-files}
# Iterate over unique chromosomes

for (Chromosome in unique(merged_top_1_percent_ROH$CHR)) {
  # Subset data for the current chromosome
  chr_data <- merged_top_1_percent_ROH[merged_top_1_percent_ROH$CHR == Chromosome, c("CHR", "POS1", "POS2")]
  
  # Define the filename with the output directory path
  filename <- file.path(output_dir_gapless_roh_hotspots, paste(prefix,"_chr", Chromosome,"_ROH_Hotspot_gapless_windows.bed", sep = "")) 
  
  # Write data to .bed file
  write.table(chr_data, file = filename, sep = "\t", quote = FALSE, col.names = FALSE, row.names = FALSE)
  
  # Print message indicating successful export
  cat("Exported", nrow(chr_data), "hotspot windows to", filename, "\n")
}

```

## 3.2: 2Mb ROH-hotspot windows
```{r Export ROH-hotpots to .bed-files}
# Iterate over unique chromosomes
for (Chromosome in unique(linked_hotspot_windows$CHR)) {
  # Subset data for the current chromosome
  chr_data <- linked_hotspot_windows[linked_hotspot_windows$CHR == Chromosome, c("CHR", "POS1", "POS2")]
  
  # Define the filename with the output directory path
  filename <- file.path(output_dir, paste(prefix,"_chr", Chromosome,"_ROH_Hotspot_windows.bed", sep = "")) 
  
  # Write data to .bed file
  write.table(chr_data, file = filename, sep = "\t", quote = FALSE, col.names = FALSE, row.names = FALSE)
  
  # Print message indicating successful export
  cat("Exported", nrow(chr_data), "hotspot windows to", filename, "\n")
}

```

